---
title: "Managing aspatial data in R"
author: "Taylor Saunders"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in WHO csv data

Use setwd command to set the working directory

```{r}
setwd("C:/Users/taylo/OneDrive/Desktop/GIS_470/Module_0")
```


```{r}
my_first_df <- read.csv("WHO.csv", head=TRUE)
```

## Exploring the dataframe

```{r}
head(my_first_df)
```
```{r}
str(my_first_df)
```
```{r}
tail(my_first_df)
```
String values/vectors: contain words, letters, or any other character type
Numerical values/vectors: hold different forms of numbers

Creating a string value:
```{r}
name_value <- "Taylor"
```

Creating a numeric value: 
```{r}
age_value <- 21
```


```{r}
print(name_value)
print(age_value)
```
```{r}
print(age_value - 10)
print(age_value/2)
```
## Vectors

Vector: a column of numeric or string values 

Creating a string vector:
```{r}
#think of c as combine
name_vector <- c("Dylan", "Sarah", "Daniel")
print(name_vector)
favorite_food <- c("Pasta", "Cheese", "Pizza")
```
Creating a numeric vector:
```{r}
age_vector <- c(30, 15, 22)
print(age_vector)
```
## Data Frames

one variable, 3 observation data frame
```{r}
df1 <- data.frame(name_vector,age_vector, favorite_food)
```


add a column/variable
```{r}
df1$age_15 <- age_vector + 15
```

```{r}
head(df1)
```


Exercise: if my age is 21, how old will I be in 2025?
```{r}
my_age <- 21
print(my_age + (2050 - 2024))
```
## Replacing DF values

Replace Sarah with Tom
```{r}
name_vector[2] <- "Tom"
print(name_vector)
```

NEW DF
```{r}
name_vector2 <- c("Dylan", "Sarah", "Lisa")
country_origin <- c("Ireland", "USA", "Mexico")
df2 <- data.frame(name_vector2, country_origin)
```


## Merging

x is the larger data frame, typically
y is the attributes we want to add into our primary df (x)
by = unique characteristic used to join the two DF

```{r}
#basic merge
df_merge <- merge(df1, df2, by.x = "name_vector", by.y = "name_vector2")
```

```{r}
head(df_merge)
```

Only Dylan and Sarah are in both vectors used to merge

Solution:
```{r}
# all = TRUE
#keep all observations
df4 <- merge(df1, df2, by.x = "name_vector", by.y = "name_vector2", all = TRUE)
```

```{r}
head(df_merge)
```

```{r}
#keep row structure of df1 and only add the columns that match in df2
df_merge <- merge(df1, df2, by.x = "name_vector", by.y = "name_vector2", all.x = TRUE)
```

```{r}
head(df_merge)
```

```{r}
df_merge <- merge(df1, df2, by.x = "name_vector", by.y = "name_vector2", all.y = TRUE)

head(df_merge)
```

## Saving as CSV
```{r}
write.csv(df_merge, "C:/Users/taylo/OneDrive/Desktop/GIS_470/Module_0/df_merge.csv")
```

reading in a csv
```{r}
df_merge_read <- read.csv("C:/Users/taylo/OneDrive/Desktop/GIS_470/Module_0/df_merge.csv", head=T)
```


## Descriptive Statistics

```{r}
head(df4)
str(df4)
```
```{r}
table(df4$name_vector)
```
duplicating the df
```{r}
#doubles
df4_dup <- rbind(df4, df4)
table(df4_dup$name_vector)
```

## Basic Statistics

```{r}
print(df4_dup)
```

```{r}
mean(df4_dup$age_vector)
```
We have missing values for Lisa
Solution:

```{r}
mean(df4_dup$age_vector, na.rm=TRUE)
```

```{r}
max(df4_dup$age_vector, na.rm=TRUE)
min(df4_dup$age_vector, na.rm=TRUE)
#standard deviation
sd(df4_dup$age_vector, na.rm=TRUE)
summary(df4_dup$age_vector)
#correlation
cor.test(df4_dup$age_vector, df4_dup$age_15)
```

