---
title: "Lecture 2"
author: "Taylor Saunders"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Managing Spatial Data

Note: `echo = FALSE` parameter can be added to the code chunk to prevent printing of the R code that generated the result.

Spatial Statistics is designed to explicitly tackle data/problems that are spatial in nature.

Downloading packages
When working with spatial data and shape files: sf (simple features for R)

```{r, warning = FALSE}
#install.packages("sf")
library("sf")
```

Functionality of sf:

```{r}
help(package = "sf")
```

## Reading in a shape file

Reading in the us_states shapefile:
similar to reading in a csv, just using a different function.
```{r}
states <- st_read("C:/Users/taylo/OneDrive/Desktop/GIS_470/Module_1/Mapping_files/Mapping_files/us_states.shp")
```

```{r}
#shows contents of states layer
View(states)
#contains a geometry field, which differentiates a shapefile from normal csv
#SPATIAL COMPONENT!
```

```{r}
head(states)
```
```{r}
str(states)
```

## Merging 

Read in States.cvs
```{r}
state_info <- read.csv("C:/Users/taylo/OneDrive/Desktop/GIS_470/Module_1/Mapping_files/Mapping_files/States.csv")
```

```{r}
head(state_info)
```
States.csv does not have a geography column, meaning R is not interpreting it as a shapefile

Merging:
```{r}
state_df <- merge(states, state_info, by = "STATE_NAME")
```

```{r}
head(state_df)
```
"Simple feature collection with 6 features and 12 fields" = interpreted as a SHAPEFILE

## Analysis

```{r}
#Summary statistics using spatial data frame state_df
#Question: How many states of each region?
table(state_df$SUB_REGION)
```
```{r}
mean(state_df$age_mean)
```
```{r}
summary(state_df)
```
Mapping Data

```{r}
plot(state_df)
```

\