---
title: "Lecture 2"
author: "Taylor Saunders"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Managing and recoding

```{r}
library(car)
library(ggplot2)

setwd("C:/Users/taylo/OneDrive/Desktop/GIS_470/Module_2")

data_1<-read.csv("state_top_half-1.csv",head=T)
data_2<-read.csv("state_bottom_half.csv",head=T)

head(data_1)
```
```{r}
head(data_2)
```

## Combining and writing states data
```{r}
state_data <- rbind(data_1, data_2)
head(state_data)
```
```{r}
write.csv(state_data,"combined_state_data.csv")
```

## Inspecting data and removing duplicates

inspect the structure of the data using str() and check the number of columns (ncol()) and
rows (nrow())

```{r}
str(state_data)
```
```{r}
ncol(state_data)
```
```{r}
nrow(state_data)
```
There are 50 states, but 51 observations
```{r}
table(state_data$state_name)
```
Michigan has a duplicate

This has previously been noted:
```{r}
table(state_data$state_name,state_data$Duplicate)
```

```{r}
new_state_data <- subset(state_data,Duplicate == "No")
table(new_state_data$state_name,new_state_data$Duplicate)
```

Turn state data into a factor variable 
Factor variables are coded with numbers, and then each number is assigned a label 
The storage is more efficient and factors are better suited for statistical analyses

```{r}
new_state_data$state_name<-factor(new_state_data$state_name)
levels(new_state_data$state_name)
head(new_state_data)
```

# Recoding

We are interested in understanding differences across states, it would be helpful if we had a variable that
indicated the region of the country for the state is in and also an indicator of whether our state is a high income or low income state


A recode is when you change the values of a variable based on its existing values

Check income:
```{r}
summary(new_state_data$family_income_per_capita)
```
The mean looks to be 34,709, the highest income is 62,707, and the lowest income is 24,344.

Create a dummy variable high_income_state which is populated only with zeros
```{r}
new_state_data$high_income_state <- 0
```

Using the ifelse() function, let’s assign all states with incomes from 25,344 to 34,709 a zero, and all states with incomes from 34,710 to 62,707 a one

```{r}
new_state_data$high_income_state <- ifelse(new_state_data$family_income_per_capita >= 34710, 1, 0)
```

If we did it correctly, it should show 25 high income states and 25 low income states.

```{r}
table(new_state_data$high_income_state)
```

In the recode function, we provide the variable to recode from, as well as the old and new values. Note: the recode() function comes from the car package

Create a new variable indicating the regions of the country:

```{r}
new_state_data$region <- new_state_data$state_name
```

Recoding strings looks a bit more complicated, but it’s really the same as recoding based on numbers. . .
i.e. new variable <- recode(old_text_variable, “c(‘text value 1’,‘text value 2’) = ‘new text
value’ ”)

```{r}

```
# ASSIGNMENT 1A


Read in the state_top_half.csv and state_bottom_half.csv into R.

```{r}
data_1<-read.csv("state_top_half-1.csv",head=T)
data_2<-read.csv("state_bottom_half.csv",head=T)
```

Use the rbind() function to join these data frames into one data frame.
```{r}
state_data <- rbind(data_1, data_2)
head(state_data)
```
Use a combination of the ifelse() and subset() function to remove the row with the state named "Michigan - Copy".

```{r}
state_data[state_data$state_name == 'Michigan - Copy',]

state_data_new <- subset(state_data, state_name!="Michigan - Copy",)
```

Use the ifelse() function to create a new variable that converts the region variable into a region_short variable. The region short variable should indicate whether a state is in the Northeast, Midwest, South, or West.
```{r}
state_data_new$state_region <- ifelse(state_data_new$state_name %in% c("Connecticut", "Maine", "Massachusetts", "New Hampshire", "Rhode Island", "Vermont", "New York", "New Jersey", "Pennsylvania"),
                                      "Northeast",
                                      ifelse(state_data_new$state_name %in% c("Illinois", "Indiana", "Iowa", "Kansas", "Michigan", "Minnesota", "Missouri", "Nebraska", "North Dakota", "Ohio", "South Dakota", "Wisconsin"),
                                             "Midwest",
                                             ifelse(state_data_new$state_name %in% c("Alabama", "Arkansas", "Delaware", "Florida", "Georgia", "Kentucky", "Louisiana", "Maryland", "District of Columbia", "Mississippi", "North Carolina", "Oklahoma", "South Carolina", "Tennessee", "Texas", "Virginia", "West Virginia"),
                                                    "South",
                                                    ifelse(state_data_new$state_name %in% c("Alaska", "Arizona", "California", "Colorado", "Hawaii", "Idaho", "Montana", "Nevada", "New Mexico", "Oregon", "Utah", "Washington", "Wyoming"),
                                                           "West",
                                                           "na"
                                                    )
                                             )
                                      )
)

head(state_data_new)

table(state_data_new$state_region)
```

Write this modified data frame to a csv file.
To complete this assignment, please upload your newly created csv.
```{r}
write.csv(state_data_new, "C:/Users/taylo/OneDrive/Desktop/GIS_470/Module_2/assignment_2A_saunders.csv")
```
